---
title: "A4: Incarceration"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
options(scipen = 999)

# load ALL necessary libraries
library(tidyverse)
library(knitr)
library(maps)
library(ggplot2)

# load dataset
prison_pop_url <- "https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-pop.csv?raw=true"
prison_pop <- read.csv(prison_pop_url, header = TRUE, stringsAsFactors = FALSE)

jail_pop_url <- "https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-jail-pop.csv?raw=true"
jail_pop <- read.csv(jail_pop_url, header = TRUE, stringsAsFactors = FALSE)

pop_per_hundred_thousand_url <- "https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-jail-rates.csv?raw=true"
pop_per_hundred_thousand <- read.csv(pop_per_hundred_thousand_url, header = TRUE, stringsAsFactors = FALSE)

per_hundred_thousand_1990_url <- "https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-prison-jail-rates-1990.csv?raw=true"
per_hundred_thousand_1990 <- read.csv(per_hundred_thousand_1990_url, header = TRUE, stringsAsFactors = FALSE)

per_hundred_thousand_1990_wa_url <- "https://github.com/melaniewalsh/Neat-Datasets/raw/main/us-prison-jail-rates-1990-WA.csv"
per_hundred_thousand_1990_wa <- read.csv(per_hundred_thousand_1990_wa_url, header = TRUE, stringsAsFactors = FALSE)

# source all R files; this runs each file from top to bottom, storing all our
# variables so that we can use them later in this index.Rmd file.
# HINT: Remember to set your working directory to the folder this file is in!
source('summary.R')
source('chart1.R')
source('chart2.R')
##source('map.R')

```

### Introduction
The problem domain for this assignment is looking at the data collected by the 
Vera Institute regarding incarceration data from county level jails and prisons.
By analyzing and visualizing the data collected, we can identify patterns and 
trends in incarceration, while tying in sociopolitical factors that affect people
of color. Looking at the data at the county level, rather than a broader state 
level, can help us gain a better understanding of how and potentially why people 
of color are incarcerated at a disproportional rate, and make clearer connections 
and conclusions of causes and consequences.

The first and second values I will be calculating are the county/states and type 
of city with the highest and lowest prison population in the most recent
year. This will provide information about the extremes of incarceration rates and
possibly help us think about why these certain counties have the highest or 
lowest prison populations. The third value I will be calculating is how
incarceration rates have changed for Black individuals in King County based on the
dataset per 100,000 people of Washington A more local look will also 
tell us how racial disparities come into play in a state like WA. The fourth and
fifth values will be the maximum jail population rate for both white and Black
individuals, showing peak jail populations for each group. This will 
demonstrate whether the disproportion is significant or not, and help us see the
differences in incarceration between a race group that is supported by systemic
and institutional systems, versus one that is constantly being oppressed by them. 

### Summary Information
From the dataset of entire prison population per county and state, I found that
the location with the lowest population from the most recently recorded year is
**`r lowest_prison_pop_info`**, and the highest population is 
**`r highest_prison_pop_info`**. From this, we can see that the counties have
different community types, with the highest coming from an urban city, and the 
lowest from a rural area. I then calculated change the maximum overall jail 
population for both white and Black individuals from all the years and counties.
Both values ended up being from the same location and year, **`r value_4_5_info `**.
The peak value for Black individuals came up to be **`r max_black_jail_pop`**,  
while the peak population for white individuals came up **`r max_black_jail_pop`**. 
To bring things back locally to King County, where some big cities such as
Seattle and Bellevue are located, I found that the change in incarceration of 
Black individuals from 1990-2016 to be a decrease of
**`r incarceration_difference_king_county`**. 

### The Dataset

**Who collected the data?**

The data is collected by the Vera Institute of Justice, a nonprofit research
and policy organization in the United States. They are dedicated to transforming
the justice system and addressing mass incarceration in the country through 
research.

**How was the data collected or generated?**

The data is separated into jail data and prison data. The jail data is mainly
collected through the U.S. Department of Justice Bureau of Justice Statistics
(BJS), along with data from state departments of correction to fill in missing federal data. Specifically, data was pulled from the National Corrections 
Reporting Program (NCRP), the Jail Populations (DCRP), the Annual Survey of 
Jails (ASJ), the Census of Jails (COJ), data collected by the 
U.S. Census Bureau and the Centers for Disease Control and Prevention (CDC),
and geographic information from the U.S. Department of Agriculture (USDA) and 
the U.S. Census Bureau

**Why was the data collected?**

The data was collected to provide insight and detail in variables related to
mass incarceration across prisons and jails in the U.S. The data is supposed
to help visualize concerning and notable trends and patterns among mass
incarceration within race groups. In order to address racial disparities in
the justice system, it's important to recognize and see the disparities as proven
in data. This will help with developing antiracist solutions and policies to 
mitigate racial bias in the justice system.

**How many observations (rows) are in your data?**

The jail population data has 153811 rows.

**How many features (columns) are in the data?** 

The jail population data has 28 columns. 


**What, if any, ethical questions or questions of power do you need to consider when working with this data?** 

Even though the data comes from reputable government sources, I think we should
still be wary of human bias in the data. Additionally, though we are trying to
gain more understanding of incarceration and race in the U.S., we should still
try to understand the context of the data before working with it. We need to
question the pre-existing power imbalances and racial biases that has influenced
the data collected, and recognize that the data will not be 100% accurate, which
is OK as long as that is acknowledged. 

**What are possible limitations or problems with this data?   (at least 200 words)**

Some possible limitations with the data is that there are a few missing values
that would help with analyzing the data and drawing connections. I think the 
completeness of a dataset can help increase reliability and representation.
Additionally,I wonder how accurate this data is since it comes from systems and 
structures that have a history of oppressing people of color. Though it may seem unlikely,
I think it's possible that the numbers may be misrepresented or altered by those
in charge to paint a different narrative. This would be extremely dangerous and
harmful when trying to enact meaningful change in the justice system. Some states
or the government may want to reduce the severity of the issue of racial 
disparities across prisons and jails, which means they may feel inclined
to present inaccurate data. Due to all the various confounding variables, I'm 
also curious as to how these datasets address them and work with them. I think
a variance in methods of handling confounding variables can also lead to some inconsistent
data, and I think with data this big and widespread, it's definitely hard to maintain
consistency. I also noticed that some racial groups, such as Latinx, are extremely
underreported in the datasets as compared to white or Black groups. 


### Trends Over Time Chart
```{r, echo = FALSE}
chart1
```

I chose to create this chart to help visualize how rates of prison population have changed over the years for each racial group. This helps
us look at the difference in numbers and how some groups have vast differences from others. The chart demonstrates how the smallest rate
of incarceration is the AAPI group, while the largest is Black individuals.
The amount that Black individuals surpasses the other racial groups is pretty big, and even the second most does not reach half of the population
for some years.

### Variable Comparison Chart
```{r, echo = FALSE}
chart2
```

For my second chart, I chose to show how total jail population compares to total jail population of individuals between 15 and 64 years old. This shows how ages below and older are less likely to be in jail and how the data is concentrated on ages in between. Because the trend is positive and linear, it shows how most of the total jail population is made of 15-64 year old individuals, providing insight on how age maybe also play a role in the likelihood of being in jail.

### Map

My map shows the quantity of prison populations among Black individuals by county and area. The variability between regions demonstrate how some counties have higher populations, while others have lower ones. Looking at this map can help us identify potential factors as to why these differences exist, such as location, geography, political and economic climate, etc. 