---
title: "A4: Incarceration"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
options(scipen = 999)

# load ALL necessary libraries
library(tidyverse)
library(knitr)
library(maps)
library(ggplot2)

# source all R files; this runs each file from top to bottom, storing all our
# variables so that we can use them later in this index.Rmd file.
# HINT: Remember to set your working directory to the folder this file is in!
source('summary.R')
source('chart1.R')
source('chart2.R')
source('map.R')

```

### Introduction
The problem domain for this assignment is looking at the data collected by the 
Vera Institute regarding incarceration data from county level jails and prisons.
By analyzing and visualizing the data collected, we can identify patterns and 
trends in incarceration, while tying in sociopolitical factors that affect people
of color. Looking at the data at the county level, rather than a broader state 
level, can help us gain a better understanding of how and potentially why people 
of color are incarcerated at a disproportional rate, and make clearer connections 
and conclusions of causes and consequences.

The first and second values I will be calculating are the county/states and type 
of city with the highest and lowest prison population in the most recent
year. This will provide information about the extremes of incarceration rates and
possibly help us think about why these certain counties have the highest or 
lowest prison populations. The third value I will be calculating is how
incarceration rates have changed for Black individuals in King County based on the
dataset per 100,000 people of Washington A more local look will also 
tell us how racial disparities come into play in a state like WA. The fourth and
fifth values will be the maximum jail population rate for both white and Black
individuals, showing peak jail populations for each group. This will 
demonstrate whether the disproportion is significant or not, and help us see the
differences in incarceration between a race group that is supported by systemic
and institutional systems, versus one that is constantly being oppressed by them. 

### Summary Information
From the dataset of entire prison population per county and state, I found that
the location with the lowest population from the most recently recorded year is
**`r lowest_prison_pop_info`**, and the highest population is 
**`r highest_prison_pop_info`**. From this, we can see that the counties have
different community types, with the highest coming from an urban city, and the 
lowest from a rural area. I then calculated change the maximum overall jail 
population for both white and Black individuals from all the years and counties.
Both values ended up being from the same location and year, **`r value_4_5_info `**.
The peak value for Black individuals came up to be **`r max_black_jail_pop`**,  
while the peak population for white individuals came up **`r max_black_jail_pop`**. 
To bring things back locally to King County, where some big cities such as
Seattle and Bellevue are located, I found that the change in incarceration of 
Black individuals from 1990-2016 to be a decrease of
**`r incarceration_difference_king_county`**. 

### The Dataset

**Who collected the data?**

The data is collected by the Vera Institute of Justice, a nonprofit research
and policy organization in the United States. They are dedicated to transforming
the justice system and addressing mass incarceration in the country through 
research.

**How was the data collected or generated?**

The data is separated into jail data and prison data. The jail data is mainly
collected through the U.S. Department of Justice Bureau of Justice Statistics
(BJS), along with data from state departments of correction to fill in missing federal data. Specifically, data was pulled from the National Corrections 
Reporting Program (NCRP), the Jail Populations (DCRP), the Annual Survey of 
Jails (ASJ), the Census of Jails (COJ), data collected by the 
U.S. Census Bureau and the Centers for Disease Control and Prevention (CDC),
and geographic information from the U.S. Department of Agriculture (USDA) and 
the U.S. Census Bureau

**Why was the data collected?**

The data was collected to provide insight and detail in variables related to
mass incarceration across prisons and jails in the U.S. The data is supposed
to help visualize concerning and notable trends and patterns among mass
incarceration within race groups. In order to address racial disparities in
the justice system, it's important to recognize and see the disparities as proven
in data. This will help with developing antiracist solutions and policies to 
mitigate racial bias in the justice system.

**How many observations (rows) are in your data?**

The jail population data has 153811 rows.

**How many features (columns) are in the data?** 

The jail population data has 28 columns. 


**What, if any, ethical questions or questions of power do you need to consider when working with this data?** 

Even though the data comes from reputable government sources, I think we should
still be wary of human bias in the data. Additionally, though we are trying to
gain more understanding of incarceration and race in the U.S., we should still
try to understand the context of the data before working with it. We need to
question the pre-existing power imbalances and racial biases that has influenced
the data collected, and recognize that the data will not be 100% accurate, which
is OK as long as that is acknowledged. 

**What are possible limitations or problems with this data?   (at least 200 words)**

Some possible limitations with the data is that there are a few missing values
that would help with analyzing the data and drawing connections. I think the 
completeness of a dataset can help increase reliability and representation.
Additionally,I wonder how accurate this data is since it comes from systems and 
structures that have a history of oppressing people of color. Though it may seem unlikely,
I think it's possible that the numbers may be misrepresented or altered by those
in charge to paint a different narrative. This would be extremely dangerous and
harmful when trying to enact meaningful change in the justice system. Some states
or the government may want to reduce the severity of the issue of racial 
disparities across prisons and jails, which means they may feel inclined
to present inaccurate data. Due to all the various confounding variables, I'm 
also curious as to how these datasets address them and work with them. I think
a variance in methods of handling confounding variables can also lead to some inconsistent
data, and I think with data this big and widespread, it's definitely hard to maintain
consistency. I also noticed that some racial groups, such as Latinx, are extremely
underreported in the datasets as compared to white or Black groups. 


### Trends Over Time Chart
```{r, echo = FALSE}
chart1
```

For first chart, I chose to visualize how prison populations per hundred thousand have changed over the years for each racial group. This helps us look at the difference in numbers and how some groups have significant differences from others. The chart demonstrates how the smallest rate of incarceration belongs to the White racial group, while the largest is Black. The chart illustrates how there is a huge gap between Black and the other racial groups. At the peak, the prison population of Black individuals reached above 7,000, while the other racial groups fall below 2,000 and peak white population rate being at around 500. This difference is very concerning, and while there has been a downward trend in the last two decades for Black population rates, racial disparities are still clearly and significantly present. 

### Variable Comparison Chart
```{r, echo = FALSE}
chart2
```

For my second chart, I chose to look at the relationship between time and Black jail population rate by hundred thousand in King County, WA from 1990-2018 (beginning to end of dataset, excluding the years where there was no data). I chose this because King County covers a vast amount of area in WA and has some big cities, so I thought this would be a meaningful relationship to visualize, especially since this is happening right around us. By looking at this chart, I can see that there has been a negative, moderately strong relationship between the years and Black jail rate in King County. As the years go on,
the population rate dropped from almost 1,500 to just above 800. I thought that this was an interesting thing to see as I did expect it to have a strong, positive relationship and that population rates have gone up. This was surprising and I think it demonstrates potential progress in racial profiling and racism in the jail system. 

### Map
```{r, echo = FALSE, fig.width=12}
map_black_2018
```

My map shows the variation across states of median jail population rate of Black individuals, pulled specifically from 2018. I chose 2018 because it was the most recent year in the dataset, and I feel that it is most timely as it is closest to 2023. I chose calculating the median of each state to measure the central tendency of Black jail population rates, and I did not choose calculating the mean as the data had more extremes where some counties in certain states had a population of zero, while others had much higher numbers that seemed to be outliers. From my map, I can see that Utah and Kentucky have the highest medians, which means that taking in account all the countries of each state, they overall had a higher Black jail population rate. It also looks like Oregon is one of the lowest, illustrated by its light state color. It does seem that the midwest, South, and East Coast have higher rates, aside from Utah. Seeing this visualization can help generate possibilites for why certain states have higher or lower rates, such as Black population, difference in policies/laws, etc.  